Clustering analysis and statistics

### 2019-08-12 notebooks/Clustering analysis.Rmd
adjusted ST trees

```{r libraries, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(ape)
library(tibble)
library(knitr)
library(vegan)
library(distanceR)
library(impoRt)
library(readxl)
library(treeio)
library(kableExtra)
```

@@

@@

@@

# Maximum likelihood trees on clades of interest
Each tree is midpoint rooted for visual purposes. The values on each node is the UF Bootstrap values calculated by IQTree. The tip point shape represents the year of isolation, and the color the animal species of origin. The tip labels represent the location of the isolate; the first number represent the county, and the second number the municipality. 
```{r message=FALSE, warning=FALSE}
# Import isolate information and tree metadata
isol_info <- read.table("../data/id.txt",
                        sep = "\t",
                        header = TRUE,
                        stringsAsFactors = FALSE)

location_data <- read_excel("../data/st_location_data.xlsx") %>%
  select(id, Kommune, Fylke) %>%
  mutate(fyl_id = group_indices(., Fylke),
         kom_id = group_indices(., Kommune)) %>%
  mutate(Location = paste(fyl_id, kom_id, sep = "-")) %>%
  select(id, Location)

@@

# Clean tree tip labels
clean_trees <- lapply(
  raw_trees,
  function(x) fix_tip_labels(
    x,
    isol_info,
    "_pilon_spades.fasta",
    tree_type = "treedata"
    )
  )

# Specify color palette
palette <- c("Broiler" = "#4575b4",
             "Pig" = "#74add1",
             "Red fox" = "#f46d43",
             "Wild bird" = "#fdae61")

# Generate trees
annotate_tree(
  clean_trees$ST10,
  tree_metadata,
  layout = "rectangular",
  tree_type = "treedata",
  midroot = TRUE,
  line_width = 0.5,
  label_variable = "Location",
  color_variable = "Species",
  shape_variable = "Year",
  clade_label_node = 22,
  clade_label = "10 SNP diff",
  cladelabel_offset = 0.00055,
  bootstrap_lab = TRUE,
  tippoint_size = 4,
  nodepoint_size = 2,
  label_offset = 0.0001,
  own_palette = palette
) +
  xlim(0, 0.005) +
  ggtitle("ST10 Clade")

annotate_tree(
  clean_trees$ST355,
  tree_metadata,
  layout = "rectangular",
  tree_type = "treedata",
  midroot = TRUE,
  line_width = 0.5,
  label_variable = "Location",
  color_variable = "Species",
  shape_variable = "Year",
  clade_label_node = 23,
  clade_label = "3 SNP diff",
  cladelabel_offset = 0.0013,
  bootstrap_lab = TRUE,
  tippoint_size = 4,
  nodepoint_size = 2,
  label_offset = 0.0003,
  own_palette = palette
) +
  xlim(0, 0.0165) +
  ggtitle("ST355 Clade")

annotate_tree(
  clean_trees$ST602,
  tree_metadata,
  layout = "rectangular",
  tree_type = "treedata",
  midroot = TRUE,
  line_width = 0.5,
  label_variable = "Location",
  color_variable = "Species",
  shape_variable = "Year",
  clade_label_node = 10,
  clade_label = "18 SNP diff",
  cladelabel_offset = 0.001,
  bootstrap_lab = TRUE,
  tippoint_size = 4,
  nodepoint_size = 2,
  label_offset = 0.0002,
  own_palette = palette
) +
  xlim(0, 0.009) +
  ggtitle("ST602 Clade")

annotate_tree(
  clean_trees$ST744,
  tree_metadata,
  layout = "rectangular",
  tree_type = "treedata",
  midroot = TRUE,
  line_width = 0.5,
  label_variable = "Location",
  color_variable = "Species",
  shape_variable = "Year",
  clade_label_node = 12,
  clade_label = "50 SNP diff",
  cladelabel_offset = 0.00024,
  bootstrap_lab = TRUE,
  tippoint_size = 4,
  nodepoint_size = 2,
  label_offset = 0.00007,
  own_palette = palette
) +
  xlim(0, 0.0030) +
  ggtitle("ST744 Clade")

>>>>

```{r libraries, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(ape)
library(tibble)
library(knitr)
library(vegan)
library(distanceR)
library(impoRt)
library(readxl)
library(treeio)
library(adephylo)
library(RColorBrewer)
library(kableExtra)
```

@@

mlst_data <- read.table("../data/mlst_results.txt",
                        sep = "\t",
                        header = TRUE,
                        stringsAsFactors = F)

@@

# Patristic distances
Patristic distances were calculated from the core gene alignment SNP tree.
```{r fig.width=8, fig.height=8, warning=FALSE, message=FALSE}
iqtree <- read.tree("../data/IQTree/core_genes_roary.treefile")
iqtree_clean <- fix_tip_labels(iqtree, id_data, "_pilon_spades.fasta", tree_type = "phylo")

pat_dist <- as.matrix(distTips(iqtree_clean, method = "patristic"))

pat_dist_bin <- as.data.frame(ifelse(pat_dist < 0.04, "1", "0")) %>%
  rownames_to_column("isol1")

row_val <- as.data.frame(pat_dist_bin$isol1) %>%
  rename("id" = `pat_dist_bin$isol1`) %>%
  left_join(mlst_data[,c("id","ST")], by = "id") %>%
  group_by(ST) %>%
  mutate(n = 1:n(),
         new_id = paste(ST, n, sep = "_"))

col_val <- as.data.frame(names(pat_dist_bin)) %>%
  rename("id" = `names(pat_dist_bin)`) %>%
  left_join(mlst_data[,c("id","ST")], by = "id") %>%
  group_by(ST) %>%
  mutate(n = 1:n(),
         new_id = paste(ST, n, sep = "_")) %>%
  ungroup() %>%
  mutate(new_id = ifelse(new_id == "NA_1", "isol1", new_id),
         ST = ifelse(is.na(ST) == TRUE, "isol1", ST))

pat_dist_clean <- pat_dist_bin
names(pat_dist_clean) <- col_val$new_id
pat_dist_clean$isol1 <- row_val$new_id

plot_data <- pat_dist_clean %>%
  gather(isol2, value, -isol1) %>%
  filter(isol1 != isol2)

ggplot(plot_data, aes(reorder(isol1, -as.numeric(value)), reorder(isol2, -as.numeric(value)), fill = value)) +
  geom_tile() +
  scale_fill_manual(values = c("1" = "#80b1d3",
                               "0" = "grey95"),
                    labels = c("Pat. dist > 0.04",
                               "Pat. dist < 0.04")) +
  labs(fill = NULL) +
  theme(axis.text = element_text(size = 3),
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.3),
        panel.background = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank()) +
  coord_fixed()
```

@@

# Maximum likelihood trees on clades of interest
Each tree is midpoint rooted for visual purposes. The values on each node is the UF Bootstrap values calculated by IQTree. The tip point shape represent the animal species of origin, and the color the year of isolation. The tip labels represent the location of the isolate; the first number represent the county, and the second number the municipality. 
```{r message=FALSE, warning=FALSE}
# Import isolate information and tree metadata
location_data <- read_excel("../data/st_location_data.xlsx") %>%
  select(id, Kommune, Fylke) %>%
  mutate(fyl_id = group_indices(., Fylke),
         kom_id = group_indices(., Kommune)) %>%
  mutate(Location = paste(fyl_id, kom_id, sep = "-")) %>%
  select(id, Location)

@@

# Clean tree tip labels
clean_trees <- lapply(
  raw_trees,
  function(x) fix_tip_labels(
    x,
    id_data,
    "_pilon_spades.fasta",
    tree_type = "treedata"
    )
  )

# Specify color and shape palettes
palette_shape <- c(
  "Broiler" = 15,
  "Pig" = 18,
  "Red fox" = 16,
  "Wild bird" = 17
)

palette_color <- brewer.pal(11, "Set3")

year_val <- c(
  "2006",
  "2007",
  "2008",
  "2009",
  "2010",
  "2011",
  "2012",
  "2013",
  "2014",
  "2015",
  "2016")

names(palette_color) <- year_val

# Generate trees
annotate_tree(
  clean_trees$ST10,
  tree_metadata,
  layout = "rectangular",
  tree_type = "treedata",
  midroot = TRUE,
  line_width = 0.5,
  label_variable = "Location",
  color_variable = "Year",
  shape_variable = "Species",
  clade_label_node = 19,
  clade_label = "10 SNP diff",
  cladelabel_offset = 0.00055,
  bootstrap_lab = TRUE,
  bootlab_size = 2,
  tippoint_size = 4,
  label_offset = 0.0001,
  shape_palette = palette_shape,
  color_palette = palette_color
) +
  xlim(0, 0.005) +
  ggtitle("ST10 Clade")

annotate_tree(
  clean_trees$ST355,
  tree_metadata,
  layout = "rectangular",
  tree_type = "treedata",
  midroot = TRUE,
  line_width = 0.5,
  label_variable = "Location",
  color_variable = "Year",
  shape_variable = "Species",
  clade_label_node = 26,
  clade_label = "6 SNP diff",
  cladelabel_offset = 0.0013,
  bootstrap_lab = TRUE,
  bootlab_size = 2,
  tippoint_size = 4,
  label_offset = 0.0003,
  shape_palette = palette_shape,
  color_palette = palette_color
) +
  xlim(0, 0.0165) +
  ggtitle("ST355 Clade")

annotate_tree(
  clean_trees$ST602,
  tree_metadata,
  layout = "rectangular",
  tree_type = "treedata",
  midroot = TRUE,
  line_width = 0.5,
  label_variable = "Location",
  color_variable = "Year",
  shape_variable = "Species",
  clade_label_node = 10,
  clade_label = "18 SNP diff",
  cladelabel_offset = 0.001,
  bootstrap_lab = TRUE,
  bootlab_size = 2,
  tippoint_size = 4,
  label_offset = 0.0002,
  shape_palette = palette_shape,
  color_palette = palette_color
) +
  xlim(0, 0.009) +
  ggtitle("ST602 Clade")

annotate_tree(
  clean_trees$ST744,
  tree_metadata,
  layout = "rectangular",
  tree_type = "treedata",
  midroot = TRUE,
  line_width = 0.5,
  label_variable = "Location",
  color_variable = "Year",
  shape_variable = "Species",
  clade_label_node = 12,
  clade_label = "50 SNP diff",
  cladelabel_offset = 0.00024,
  bootstrap_lab = TRUE,
  bootlab_size = 2,
  tippoint_size = 4,
  label_offset = 0.00007,
  shape_palette = palette_shape,
  color_palette = palette_color
) +
  xlim(0, 0.0030) +
  ggtitle("ST744 Clade")
